---
title: Daily Nest Survival Modeling of Wild Turkeys in the Mid-Atlantic Region
authors: "K. Smelter, F. Buderman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
    toc: true
---
 
**Purpose**: This script uses derived incubation start and end dates to fit a Bayesian known fate model 
**Key Changes**: This script uses the cloglog link instead of the logit link for modeling daily nest survival
**Last Updated**: 4/27/25

```{r}
################################################################################
## Load Packages

library(nimble)
library(MCMCvis)
library(tidyverse)
library(sf)
library(GGally)
library(matrixStats)
library(tidyverse)
library(tidybayes)
library(VGAM)
library(ggpubr)

load("Data Management/RData/Known Fate/Pennsylvania/Manuscript/01_Covs_Ready.RData")


################################################################################
## Data Prep- Encounter Histories
```

Convert start and end of incubation days to date objects

```{r}
nests.scaled$startI <- as.Date(nests.scaled$startI)
nests.scaled$endI <- as.Date(nests.scaled$endI)
```

Create a sequence of all unique dates between the min start and max end date

```{r}
nests.scaled %>% 
  sf::st_drop_geometry() %>%
  group_by(NestYr) %>% summarise(mindate = min(startI),
                                                maxdate = max(endI))
```

Get minimum and maximum days within a year

```{r}
min(nests.scaled$startI)
max(nests.scaled$endI)
inc.dates <- sort(unique(c(nests.scaled$startI, nests.scaled$endI)))
inc.dates <- seq(as.Date("2022-04-19"), as.Date("2022-09-16"), by = 1)
inc.dates <- format(inc.dates, "%m-%d")
inc.dates
```

Calculate the number of encounter days

```{r}
occs <- length(inc.dates)
```

Initialize variables
Create matrix for encounter histories

```{r}
first <- last <- c()
surv.caps <- matrix(data = NA, nrow = nrow(nests.scaled), ncol = occs)
```

Create encounter histories within surv.caps
If a nest fails it gets a zero at the end 
If it reaches hatch it ends with a 1

```{r}
for(i in 1:nrow(nests.scaled)){ 
  first[i] <- which(inc.dates == format(nests.scaled$startI[i], "%m-%d")) 
  last[i] <- which(inc.dates == format(nests.scaled$endI[i], "%m-%d")) 
  surv.caps[i,first[i]:last[i]] <- 1 
  if(nests.scaled$NestFate[i] == 0)
  {surv.caps[i,last[i]] <- 0} 
}
```

Check work on first individual

```{r}
surv.caps[11,]
first;
last;
```

Change surv.caps to a df

```{r}
surv.caps = as.data.frame(surv.caps)
colnames(surv.caps) <- inc.dates
surv.caps$Year = nests.scaled$NestYr-2021


################################################################################
## Bring in Weather Data
```

Functions from Kery and Schaub to get the first and last encounter occasions
Make sure that f is always greater than k (First encounter always comes before last encounter)

```{r}
getFirst <- function(x) {min(which(!is.na(x)))}
getLast <- function(x) {max(which(!is.na(x)))}
f <- apply(surv.caps[,1:151],1,getFirst)
k <- apply(surv.caps[,1:151],1,getLast)
f;k
f<k
```

Store weather data within arrays for 2022 and 2023
Rows = 158 (# of nests)
Columns = 365 (# of days within a year)
Dimensions = 1, or 2 (Temp = 1, Precip = 2)

```{r}
temperatureC2022 <- weather.array[1:158,1:365,1]
temperatureC2023 <- weather.array[1:158,366:730,1]
precip2022 <- weather.array[1:158,1:365,2]
precip2023 <- weather.array[1:158,366:730,2]

################################################################################
## Get 3 day moving average
```

Convert from matrix to dataframe
For each row in surv.caps if the year is the first year of the study, collect the rolling mean for 2022 variables
For each row in surv.caps if the year is the second year of the study, collect the rolling mean for 2023 variables

```{r}
testTempC3Roll = data.frame(matrix(NA, nrow = 158, ncol = 151))   
testPrecip3Roll = data.frame(matrix(NA, nrow = 158, ncol = 151))
for(i in 1:nrow(surv.caps)){
  if(surv.caps$Year[i] == 1){
    for(j in (f[i]+108):(k[i]+108)){
       testTempC3Roll[i,j-108] <- mean(temperatureC2022[i,((j-2):j)])
       testPrecip3Roll[i,j-108] <- mean(precip2022[i,((j-2):j)])
     }
   }
   else if(surv.caps$Year[i] == 2){
      for(j in (f[i]+108):(k[i]+108)){
     testTempC3Roll[i,j-108] <- mean(temperatureC2023[i,((j-2):j)])
    testPrecip3Roll[i,j-108] <- mean(precip2023[i,((j-2):j)])
    }
  }
   else{
  testTempC3Roll[i,j] <- "You messed up"
  testPrecip3Roll[i,] <- "You messed up"
}
 }
```

Check work

```{r}
range(testPrecip3Roll, na.rm = T) 
range(testTempC3Roll, na.rm = T)
```

Scale weather data
Scale function had problems with the NAs within the matrices

```{r}
weather.array2 <- array(data = NA, dim = c(nrow(testTempC3Roll), ncol(testTempC3Roll),2))
testTempC.mat <- as.matrix(testTempC3Roll)
testTempC.scale <- (testTempC.mat - mean(testTempC.mat,na.rm=T))/sd(testTempC.mat,na.rm=T)
testTempC.scale
testPrecip.mat <- as.matrix(testPrecip3Roll)
testPrecip.scale <- (testPrecip.mat - mean(testPrecip.mat,na.rm=T))/sd(testPrecip.mat,na.rm=T)
testPrecip.scale
weather.array2[,,1] <- testTempC.scale
weather.array2[,,2] <- testPrecip.scale
```

Check work

```{r}
weather.array2[1,1:100,1]
weather.array2[1,1:100,2]
```

Save output RDS files

```{r}
saveRDS(weather.array2, "Known Fate Model/Pennsylvania/Weather Arrays/weather.pa_2022_2023_3daymean.RDS")
```

Check correlations

```{r}
temp_flat <- as.vector(weather.array2[,,1])
precip_flat <- as.vector(weather.array2[,,2])
```

Compute correlation, excluding NA pairs

```{r}
cor_result <- cor(temp_flat, precip_flat, use = "complete.obs")
```

Print correlation

```{r}
cor_result


################################################################################
## Compile Parameters into matrix format
```

Function to fill NA values with 0 throughout dataframe
These parameters are scaled so we are replacing NA values with the mean

```{r}
fill_NA_with_value <- function(df, value = 0) {
  df[] <- lapply(df, function(col) {
    if (is.numeric(col)) {
      replace(col, is.na(col), value)
    } else {
      col
    }
  })
  return(df)
}
nests.ready <- fill_NA_with_value(nests.scaled)
summary(nests.ready)


################################################################################
## Nimble Model

knownfate.habitat <- nimbleCode({
  
  #### Loop over individuals 
  for (i in 1:n.ind) {
    
    #### Get the first and last days a nest was active
    for(t in (first[i]+1):last[i]) {
      
      #### mu and z change if an animal is alive or dead 
      mu[i,t] <- phi[i,t]*z[i,t-1]
      
      #### Estimating daily nest survival in matrix format
      cloglog(phi[i,t]) <- inprod(beta[1:J], X[i, 1:J]) +
        weather.beta[1]*weather.array[i,t,1] + weather.beta[2]*weather.array[i,t,2] + 
        weather.beta[3]*weather.array[i,t,1]*weather.array[i,t,2]
      
      #### Likelihood of Nest Survival (Bernoulli)
      z[i,t]~dbern(mu[i, t])
      
    }
  }
  
  #### Priors for beta coefficients
  for (j in 1:J){
    beta[j] ~ dnorm(0, sd = 1)  
  }
  
  #### Priors for weather coefficients
  for (k in 1:n.weather.cov) {
    weather.beta[k] ~ dnorm(0, sd = 1)
  }
  
})

################################################################################
```

Model parameters

```{r}
X <- cbind(
  rep(1, nrow(nests.ready)),               # Intercept (1)
  nests.ready$AvgVO,                       # Visual Obstruction
  nests.ready$StemCount,                   # Woody Stem Count
  nests.ready$Basal,                       # Basal Area
  nests.ready$PercWoody,                   # Percent Woody Vegetation
  nests.ready$PercGrassForb,               # Percent Grass Forb
  nests.ready$Primary,                     # Distance to Primary Road
  nests.ready$Secondary,                   # Distance to Secondary Road
  nests.ready$Mixed,                       # Mixed Forest
  nests.ready$Evergreen,                   # Evergreen Forest
  nests.ready$Developed,                   # Developed
  nests.ready$Pasture,                     # Pasture
  nests.ready$Crop,                        # Crop
  nests.ready$Grassland,                   # Grassland
  nests.ready$Wetland,                     # Wetland
  nests.ready$PercFern,                    # Percent Fern
  nests.ready$`Nest Incubation Date`,      # Nest Incubation Date
  nests.ready$IncubationConstancy,         # Incubation Constancy
  nests.ready$age                          # Age Class
) 
```

Use ggpairs to visualize correlations

```{r}
ggpairs(as.data.frame(X), 
        upper = list(continuous = "cor"),
        diag = list(continuous = "barDiag"),
        lower = list(continuous = "points"))
```

Constants

```{r}
Consts <- list(
  n.ind = nrow(nests.ready),   
  X = X,                       
  J = ncol(X),                 
  first = first,               
  last = last,  
  n.weather.cov = dim(weather.array2)[3]+1,
  weather.array = weather.array2
)
surv.caps.matrix <- as.matrix(surv.caps)
colnames(surv.caps.matrix) <- NULL
```

Data (survival data: 1 = survived, 0 = failed)

```{r}
Data <- list(
  z = surv.caps.matrix               
)
```

Initial values for parameters

```{r}
Inits <- list(
  beta = rep(0, Consts$J),
  weather.beta = rep(0,Consts$n.weather.cov)
)

start <- Sys.time()
```

Run MCMC sampling with Nimble

```{r}
nimbleMCMC_samples <- nimbleMCMC(
  code = knownfate.habitat,          
  constants = Consts,          
  inits = Inits,              
  data = Data,                 
  nburnin = 3000,             
  niter = 20000,               
  thin = 1                     
)

summary(nimbleMCMC_samples)
colMeans(nimbleMCMC_samples[,1:22])
colSds(nimbleMCMC_samples[,1:22])
MCMCtrace(nimbleMCMC_samples, pdf = FALSE)
```

Extract the posterior samples for the 'beta' parameters (columns 219 to 230)

```{r}
beta_samples <- nimbleMCMC_samples[, 1:22]

# Convert mcmc.list to matrix
samples_matrix <- as.matrix(beta_samples)

# Convert the matrix to a data frame
samples_df <- as.data.frame(samples_matrix)
```

Create a vector of new names

```{r}
new_names <- c("Intercept", 
               "Visual Obstruction",
               "Woody Stem Count",
               "Basal Area",
               "Percent Woody Vegetation",
               "Percent Grass/Forb",
               "Distance to Primary Road",
               "Distance to Secondary Road",
               "Mixed Forest",
               "Evergreen Forest",
               "Developed",
               "Pasture",
               "Crop",
               "Grassland/Shrub",
               "Wetland",
               "Percent Fern",
               "Nest Incubation Date",
               "Incubation Constancy",
               "Age Class",
               "Minimum Temperature",
               "Precipitation",
               "Precipitation * Minimum Temperature")
```

Assign the variable names to the columns of the beta_samples

```{r}
colnames(samples_df) <- new_names
```

View the renamed data frame

```{r}
head(samples_df)


################################################################################
## Data Prep for Beta Plot
```

Reshape the data into long format for ggplot

```{r}
samples_long <- samples_df %>%
  pivot_longer(cols = everything(), 
               names_to = "parameter", 
               values_to = "estimate") 
```

View the reshaped data

```{r}
head(samples_long)

credible_intervals <- samples_long %>%
  group_by(parameter) %>%
  summarise(
    lower = quantile(estimate, 0.05),   
    upper = quantile(estimate, 0.95),   
    .groups = 'drop'
  )
```

Calculate posterior mean for each parameter
Calculate 95% credible intervals using quantiles

```{r}
mean_estimates <- samples_long %>%
  dplyr::group_by(parameter) %>%
  summarise(
    mean_estimate = mean(estimate),  
    lower = quantile(estimate, 0.05),  
    upper = quantile(estimate, 0.95),  
    .groups = 'drop'
  ) %>%
  dplyr::filter(parameter != "Intercept")
mean_estimates

mean_estimates <- mean_estimates %>%
  mutate(Scale = case_when(
    parameter %in% c("Percent Grass/Forb", 
                     "Percent Woody Vegetation", 
                     "Visual Obstruction",
                     "Woody Stem Count",
                     "Basal Area",
                     "Percent Fern") ~ "Nest",
    parameter %in% c( "Grassland/Shrub",
                      "Mixed Forest",
                      "Evergreen Forest",
                      "Developed",
                      "Distance to Primary Road",
                      "Distance to Secondary Road",
                      "Pasture",
                      "Crop",
                      "Wetland"
                      ) ~ "Landscape",
    parameter %in% c("Minimum Temperature",
                    "Precipitation",
                    "Precipitation * Minimum Temperature"
                    ) ~ "Weather",
    TRUE ~ "Individual"
  ))

mean_estimates
```

Organize variables into levels to be displayed
Filter out the intercept

```{r}
mean_estimates <- mean_estimates %>%
  dplyr::mutate(parameter = factor(parameter, 
                                   levels = c( 
                                     "Precipitation * Minimum Temperature",
                                     "Minimum Temperature",
                                     "Precipitation",
                                     "Wetland",
                                     "Grassland/Shrub",
                                     "Mixed Forest",
                                     "Evergreen Forest",
                                     "Developed",
                                     "Pasture",
                                     "Crop",
                                     "Distance to Primary Road",
                                     "Distance to Secondary Road",
                                     "Percent Fern",
                                     "Percent Grass/Forb",
                                     "Percent Woody Vegetation",
                                     "Visual Obstruction",
                                     "Basal Area",
                                     "Woody Stem Count",
                                     "Incubation Constancy",
                                     "Nest Incubation Date",
                                     "Age Class"
                                     )))
mean_estimates

mean_estimates <- mean_estimates %>%
  mutate(Scale = factor(Scale, levels = c("Individual", "Nest", "Landscape", "Weather")))


################################################################################
## Beta Plot 
```

Beta estimates and associated 90% credible intervals 

```{r}
p1.betas <- ggplot(mean_estimates, 
                   aes(x = parameter, 
                       y = mean_estimate, 
                       color = Scale, 
                       shape = Scale)) +
  geom_point(size = 3.5) +  
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, size = 1.1) +  
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  
  labs(x = "Parameter", y = "Posterior Mean") +
  theme_minimal() + 
  coord_flip()+
  scale_color_manual(values = c("Individual" = "#DA6509",
                                "Nest" = "#A44200",
                                "Landscape" = "#D65F5F",
                                "Weather" = "#197278")) +  
  scale_shape_manual(values = c("Individual" = 15, 
                                "Nest" = 17,
                                "Landscape" = 16,
                                "Weather" = 18)) +
  theme(
    axis.title.x = element_text(size = 14, margin = margin(t = 10), hjust = 0.45),  
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 14),
    legend.position = "none"
  )
p1.betas
```

Beta estimates and associated 90% credible intervals 

```{r}
p1.betas <- ggplot(mean_estimates, 
                   aes(x = parameter, 
                       y = mean_estimate, 
                       color = Scale, 
                       shape = Scale)) +
  geom_point(size = 3.5) +  
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, size = 1.1) +  
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  
  labs(x = "Parameter", y = "Posterior Mean") +
  theme_minimal() + 
  coord_flip()+
  scale_color_manual(values = c("Individual" = "#DA6509",
                                "Nest" = "#A44200",
                                "Landscape" = "#D65F5F",
                                "Weather" = "#197278")) +  
  scale_shape_manual(values = c("Individual" = 15, 
                                "Nest" = 17,
                                "Landscape" = 16,
                                "Weather" = 18)) +
  theme(
    axis.title.x = element_text(margin = margin(t = 10), hjust = 0.4),
    axis.title.y = element_text(margin = margin(r = 12))
    
  )
p1.betas
```

Save multiple objects in a single RData file

```{r}
 save(p1.betas, mean_estimates, samples_df, file = "MAWTRC Nesting Ecology Manuscript/Figures/RData/Known Fate/kf.betas.habitat.PA.RData", overwrite = T)


################################################################################
## DNS Plots
```

For each prediction plot update the covariate within colnames
Include the intercept and multiply the beta by the predictor
Create plots below by updating the x-axis with each predictor

```{r}
sim_tbl <- expand_grid(-1:1)
colnames(sim_tbl) <- c("Percent Fern") 

sim_tbl$prediction <- pmap(list(sim_tbl$`Percent Fern`), function(cov1){
                             clogloglink(samples_df$Intercept + 
                                         samples_df$`Percent Fern` * cov1, 
                                         inverse = TRUE)
                           })
sim_tbl_long <- unnest(sim_tbl, prediction) 

pred.p1 <- sim_tbl_long %>%
  ggplot(aes(x = `Percent Fern`, y = prediction)) +
  stat_lineribbon(.width = c(.95, .8, .5), alpha = 0.8, point_interval = mean_qi) +
  scale_fill_manual(values = rev(c("#636363", "#bdbdbd", "#f0f0f0"))) +
  ylab("Daily Nest Survival Probability") +
  theme_minimal() +
  theme(
    axis.title.y = element_text(margin = margin(r = 15)),
    axis.title.x = element_text(margin = margin(t = 15))
     )
pred.p1

pred.p2 <- sim_tbl_long %>%
  ggplot(aes(x = `Incubation Constancy`, y = prediction)) +
  stat_lineribbon(.width = c(.95, .8, .5), alpha = 0.8, point_interval = mean_qi) +
  scale_fill_manual(values = rev(c("#636363", "#bdbdbd", "#f0f0f0"))) +
  ylab("Daily Nest Survival Probability") +
  theme_minimal() +
  theme(
    axis.title.y = element_text(margin = margin(t = 15)),
    axis.title.x = element_text(margin = margin(t = 15))
  )
pred.p2

################################################################################
###############################################################################X
```

